---
title: "201910_SIGN_STAT2 GSEA normalized by goseq"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(collapse=TRUE, comment="#>", warning=FALSE, message=FALSE)
```

# GSEA

This GSEA benchmarking will perform GSEA on the wild type samples from the
`201910_SIGN_STAT2` data for both GO terms and KEGG pathways using the
`EnrichmentBrowser` package.

*Note:* `goseq` normalizes raw counts for gene length *after* DE analysis

```{r load_libraries, message=FALSE, warning=FALSE, echo=FALSE}
library("EnrichmentBrowser")
library("SummarizedExperiment")
library("dplyr")
library("goseq")
library("org.Mm.eg.db")
```

Results data from *prior DE analysis.* The results columns are selected, given
appropriate column names and converted to a `SummarizedExperiment` object.

```{r}
wt_counts_res <- read.csv("./data/mouse_results_1.csv", header=TRUE, sep=",") %>% 
                 # rename(FC = log2FoldChange) %>%
                 # rename(DESeq2.STAT = stat) %>%
                 # rename(PVAL = pvalue) %>%
                 # rename(ADJ.PVAL = padj) %>%
                 dplyr::select(Gene:padj)
# names(res1df) <- c("Gene", "geneName", "baseMean", "FC", "DESeq2.STAT", "PVAL", "ADJ.PVAL")
# rownames(wt_counts_res) <- wt_counts_res$Gene
# wt_counts_res <- res1 %>% dplyr::select(-Gene)

wt_counts_res
```

Next, we format the DE genes into a vector suitable for use with `goseq`

```{r}
wt_counts_res <- wt_counts_res %>% mutate(binary = case_when(
                 (wt_counts_res$padj < 0.05 & !is.na(wt_counts_res$padj)) ~ 1,
                 (wt_counts_res$padj < 0.05 & is.na(wt_counts_res$padj)) ~ 0,
                 (is.na(wt_counts_res$padj)) ~ 0))

DE_genes <- as.integer(wt_counts_res$binary)
DE_genes[is.na(DE_genes)] <- 0
names(DE_genes) <- wt_counts_res$Gene
table(DE_genes)
```

### Check for organism info availability

```{r}
# View(supportedOrganisms())
# supportedOrganisms()[supportedOrganisms()$Genome=="mm9",]
```

Read `featurecounts` gene file to get gene lengths for every gene. 

```{r}
gene_len <- read.table("./data/STAT2-TPOS1_featurecounts.tsv", skip = 1,
                       sep = '\t', header = TRUE)
names(wt_counts_res)[1] <- "Geneid"
gene_table <- left_join(wt_counts_res, gene_len, by=c("Geneid"))
gene_len <- gene_table %>% filter(!is.na(length))
# idx <- match(gene_len$Geneid, wt_counts_res$Gene)
# gene_len$selected <- gene_len$Geneid[idx]
gene_length <- as.integer(gene_len$Length)
names(gene_length) <- gene_len$Geneid
head(gene_length)
```

### Fitting the Probability Weighting Function (PWF).

We first need to obtain a weighting for each gene, depending on its length,
given by the PWF. `nullp` plots the resulting fit, allowing verification of the
goodness of fit before continuing the analysis. Further plotting of the PWF can
be performed using the `plotPWFfunction`.

```{r}
pwf=nullp(DE_genes, bias.data=gene_length)
```

```{r}
head(pwf)
```


### Using the Wallenius approximation

To start with we will use the default method, to calculate the over and under
expressed GO categories among DE genes. Again, we allow `goseq` to fetch data
automatically, except this time the data being fetched is the relationship
between ENSEMBL gene IDs and GO categories.

```{r}
GO.wall=goseq(pwf, "mm9", "ensGene") # need to provide mm10 GO terms somehow
head(GO.wall)
```

```{r}
# library("org.Mm.egENSEMBL2EG")
# library("org.Mm.egPATH")
# Get the mapping from ENSEMBL 2 Entrez
en2eg=as.list(org.Mm.egENSEMBL2EG)
# Get the mapping from Entrez 2 KEGG
eg2kegg=as.list(org.Mm.egPATH)
# Define a function which gets all unique KEGG IDs
# associated with a set of Entrez IDs
grepKEGG=function(id,mapkeys){unique(unlist(mapkeys[id], use.names=FALSE))}
# Apply this function to every entry in the mapping from
# ENSEMBL 2 Entrez to combine the two maps
kegg=lapply(en2eg, grepKEGG, eg2kegg)
head(kegg)
```


```{r}
KEGG=goseq(pwf, gene2cat=kegg)
head(KEGG)
```

First we have to get the correct *KEGG* pathway names for the `goseq` results.

```{r}
# https://biobeat.wordpress.com/category/r/
# http://rest.kegg.jp/list/pathway/
getKeggPathwayTable <- function(){
  pathway_link_REST_url <- "http://rest.kegg.jp/list/pathway/"
  kegg_pathways<- data.frame()
 
  current_row = 1
  for (line in readLines(pathway_link_REST_url)) {
    tmp <- strsplit(line, "\t")[[1]]
    map <- tmp[1]
    map <- strsplit(map, ":")[[1]][2]  
    pathway_name<- tmp[2]
    kegg_pathways[current_row, 1]=map
    kegg_pathways[current_row, 2]=pathway_name
    current_row = current_row + 1
  }
  names(kegg_pathways) <- c("map_id","name")
  kegg_pathways
}

kegg_pathways <- getKeggPathwayTable()
```

```{r}
## clean up the goseq terms to match the edaseq go results
kegg_pathways$category <- gsub("map", "", kegg_pathways$map_id)
kegg_mapped <- full_join(KEGG, kegg_pathways, by = "category")
kegg_mapped$mmu <- gsub("map", "mmu", kegg_mapped$map_id) 
kegg_mapped$term <- paste0(kegg_mapped$mmu, "_", kegg_mapped$name)
kegg_mapped$term <- gsub(" ", "_", kegg_goseq_norm_term)
kegg_mapped <- kegg_mapped %>%
  dplyr::select(term, category, over_represented_pvalue, under_represented_pvalue)
kegg_mapped
```

```{r}
#### Save Results
saveRDS(GO.wall, file = "./results/go_goseq_mouse.rds")
saveRDS(kegg_mapped, file = "./results/kegg_goseq_mouse.rds")
```

*Session Info*

```{r sessionInfo}
sessionInfo()
```