---
title: "GSEA on WT counts (not normalized for length)"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(collapse=TRUE, comment="#>", warning=FALSE, message=FALSE)
```

# GSEA

This GSEA benchmarking will perform GSEA on the wild type samples from the
`rms_sk_combo` data for both GO terms and KEGG pathways using the
`EnrichmentBrowser` package.

```{r load_libraries, message=FALSE, warning=FALSE, echo=FALSE}
library("EnrichmentBrowser")
library("SummarizedExperiment")
library("dplyr")
library("readr")
```

Raw counts from `rms_sk_combo` are read from a *.csv* file.

```{r}
counts <- read.csv("./data/rms_sk_combo.csv", header=TRUE, sep=",")
remove_after_dot <- counts$X
names(counts)[1] <-"Gene"
counts$Gene <- gsub("\\..*","", remove_after_dot)
counts <- counts %>% dplyr::select(Gene, T01:T03, KMR19, KMR46)
rownames(counts) <- counts$Gene
counts <- counts %>% dplyr::select(-Gene)
counts
```

Column data from `rms_sk_combo` is read and prepared.

```{r}
## Assign condition 'coldata' 
condition <- factor(c(rep("T0", 3), rep("KMR", 2)))
coldata <- data.frame(row.names=colnames(counts), condition)
```

The raw counts and column data are combined in a `SummarizedExperiment` object.

```{r}
counts <- as.matrix(as.data.frame(counts))
mouse_counts_se <- SummarizedExperiment(assays=list(counts=counts),
                                     colData=coldata)
head(assay(mouse_counts_se))
```

Add the DE results to the `rowData` of the `SummarizedExperiment`

```{r deana, message=FALSE, warning=FALSE}
res1 <- read.csv("./data/mouse_results_1.csv", header=TRUE, sep=",") # %>% 
                 # rename(FC = log2FoldChange) %>% 
                 # rename(DESeq2.STAT = stat) %>% 
                 # rename(PVAL = pvalue) %>%
                 # rename(ADJ.PVAL = padj) %>%
                 # dplyr::select(Gene, FC, DESeq2.STAT, PVAL, ADJ.PVAL)
res1 <- res1 %>% dplyr::select(Gene:padj)
names(res1) <- c("Gene", "GeneName", "baseMean", "FC", "lfcSE", "DESeq2.STAT",
                 "PVAL", "ADJ.PVAL")
# Add gene name column 1
idx <- match(res1$Gene, rownames(counts))
res1$selected <- res1$Gene[idx]
res1 <- res1[complete.cases(res1$selected), ]
res1 <- res1 %>% dplyr::select(FC, DESeq2.STAT, PVAL, ADJ.PVAL)
# dim(res1)

# rownames(res1) <- res1$Gene
# countdata_res <- res1 %>% dplyr::select(-Gene)
# counts_res <- as.matrix(as.data.frame(countdata_res))
# colData_res <- names(countdata_res)
# mouse_res_se <- SummarizedExperiment(assays=list(counts=counts_res),
#                                          colData=colData_res)
 
rowData(mouse_counts_se, use.names = TRUE) <- res1
rowData(mouse_counts_se)
```

#### ID mapping

Using genomic information from different resources often requires mapping
between different types of gene identifiers. Although primary analysis steps
such as normalization and differential expression analysis can be carried out
independent of the gene ID type, downstream exploration functionality of the
`EnrichmentBrowser` is consistently based on *NCBI Entrez Gene IDs*. It is
beneficial to initially map gene IDs of a different type to *NCBI Entrez IDs*.
The function `idTypes` lists the available ID types for the mapping depending on
the organism under investigation.

```{r idtypes, message=FALSE, warning=FALSE}
idTypes("mmu")
```

ID mapping for the `airway` dataset (from *ENSEMBL* to *ENTREZ* gene ids) can
then be carried out using the function `idMap`.

```{r}
head(rownames(mouse_counts_se))
```

*ENSEMBL IDs* are mapped to the *ENTREZ IDs*, and the `airway` gene expression
data is ready to be subjected to enrichment analysis.

```{r idmap, message=FALSE, warning=FALSE}
mouse_counts_se <- idMap(mouse_counts_se, org="mmu", from="ENSEMBL", to="ENTREZID")
head(rownames(mouse_counts_se))
```

### Enrichment Analysis

In the following, we introduce how the `EnrichmentBrowser` package can be used
to perform state-of-the-art enrichment analysis of gene sets. We consider the
`airway` gene expression data as processed in the previous sections. We are now
interested in whether predefined sets of genes that are known to work together,
e.g. as defined in the *Gene Ontology (GO)* or the *KEGG* pathway annotation,
are coordinately differentially expressed. The function `getGenesets` can be
used to download gene sets from databases such as *GO* and *KEGG*. Here, we use
the function to download all *KEGG* pathways for a chosen organism (here:
*Homo sapiens*) as gene sets. Analogously, the function `getGenesets` can also
be used to retrieve *GO* terms of a selected ontology (such as biological
process, *BP*) as defined in the `GO.db` annotation package.

```{r keggs, message=FALSE, warning=FALSE}
## retrieve KEGG pathway terms of a selected organism
## https://rdrr.io/bioc/EnrichmentBrowser/man/getGenesets.html
kegg.gs <- getGenesets(org="mmu", db="kegg")
## retrieve GO terms of a selected organism
go.gs <- getGenesets(org="mmu", db="go", go.mode="GO.db")
```

#### Set Based Enrichment Analysis (SBEA)

Set-based enrichment analysis (SBEA). This is the main function for the
enrichment analysis of gene sets. It implements and wraps existing
implementations of several frequently used methods and allows a flexible
inspection of resulting gene set rankings. Currently, the following set-based
enrichment analysis methods are supported by the function `sbeaMethods()`:
 "ora", "safe", "gsea", "gsa", "padog", "globaltest", "roast", "camera", "gsva",
"samgs", "ebm" and "mgsa".

The method used here within the `sbea()` function is
*GSEA (Gene Set Enrichment Analysis)* which is frequently used and widely
accepted. *GSEA* uses a Kolmogorov–Smirnov statistic to test whether the ranks
of the p-values of genes in a gene set resemble a uniform distribution. An *α*
parameter value of 0.05 and 1000 permutations will be used.

NOTE:: The Kolmogorov–Smirnov test is a nonparametric test of the equality of
continuous or discontinuous, one-dimensional probability distributions that can
be used to compare a sample with a reference probability distribution. The
Kolmogorov–Smirnov statistic quantifies a distance between the empirical
distribution function of the sample and the cumulative distribution function of
the reference distribution, or between the empirical distribution functions of
two samples. The null distribution of this statistic is calculated under the
null hypothesis that the sample is drawn from the reference distribution (in the
one-sample case) or that the samples are drawn from the same distribution (in
the two-sample case). In the one-sample case, the distribution considered under
the null hypothesis may be continuous, purely discrete or mixed. In the
two-sample case, the distribution considered under the null hypothesis is a
continuous distribution but is otherwise unrestricted. 

The result of every enrichment analysis is a ranking of gene sets by the
corresponding p-value. The `gsRanking` function displays only those gene sets
satisfying the chosen significance level *α*.

```{r}
gsea_res <- sbea(method="gsea", se=mouse_counts_se, gs=go.gs, alpha = 0.05,
                 perm = 1000)
gsRanking(gsea_res) # DataFrame with gene sets ranked by the corresponding p-val
```

### Network-based enrichment analysis

We are now interested whether these findings can be supported by known
regulatory interactions. For example, we want to know whether transcription 
factors and their target genes are expressed in accordance to the connecting
regulations (activation/inhibition). Such information is usually given in a gene
regulatory network derived from specific experiments or compiled from the
literature. Here, we compile a network from regulations in pathway databases
such as *KEGG*. Now, we are able to perform enrichment analysis using the
compiled network.

```{r}
## Compilation of a gene regulatory network from pathway databases
mmu.grn <- compileGRN(org="mmu", db="kegg")
head(mmu.grn)
```

Network-based enrichment analysis (NBEA). This is the main function for
network-based enrichment analysis. It implements and wraps existing
implementations of several frequently used methods and allows a flexible
inspection of resulting gene set rankings. Currently, the following set-based
enrichment analysis methods are supported by the function `nbeaMethods()`:
"ggea", "spia", "pathnet", "degraph", "ganpa", "cepa", "topologygsa" and
"netgsa".

The method used here within the `nbea()` function is
*GGEA: Gene Graph Enrichment Analysis*  which evaluates the consistency of known
regulatory interactions with the observed expression data. We perform *GGEA*
using the compiled *KEGG* regulatory network.

```{r}
ggea_res <- nbea(method="ggea", se=mouse_counts_se, gs=go.gs, grn=mmu.grn,
                 alpha = 0.05)
gsRanking(ggea_res)
```

The resulting ranking lists, for each statistically significant gene set, the
number of relations of the network involving a member of the gene set under
study *(NR.RELS)*, the sum of consistencies over the relations of the set
*(RAW.SCORE)*, the score normalized by induced network size
*(NORM.SCORE = RAW.SCORE / NR.RELS)*, and the statistical significance of each
gene set based on a permutation approach.

<!-- A *GGEA* graph for a gene set depicts -->
<!-- the consistency of each interaction in the set. Nodes (genes) are colored -->
<!-- according to expression (up-/down-regulated) and edges (interactions) -->
<!-- are colored according to consistency, i.e. how well the interaction type -->
<!-- (activation/inhibition) is reflected in the correlation of the observed -->
<!-- expression of both interaction partners. -->

<!-- ```{r fig.height=6, fig.width=12} -->
<!-- par(mfrow=c(1,2)) -->
<!-- ggeaGraph(gs=mmu.gs[["0001666_response_to_hypoxia"]], -->
<!--           grn=mmu.grn, se=wt_counts_se) -->
<!-- ggeaGraphLegend() -->
<!-- ``` -->

#### Combining results

Different enrichment analysis methods usually result in different gene set
rankings for the same dataset. To compare results and detect gene sets that are
supported by different methods, the `EnrichmentBrowser` package allows to
combine results from the different set-based and network-based enrichment
analysis methods. The combination of results yields a new ranking of the gene
sets under investigation by specified ranking criteria, e.g. the average rank
across methods. We consider the *GSEA* result and the *GGEA* result from the
previous sections and use the function `combResults`.

```{r}
res.list <- list(gsea_res, ggea_res)
comb.res <- combResults(res.list)
```

The combined result can be detailedly inspected as before and interactively
ranked. Using the `eaBrowse` function creates a HTML summary from which each
gene set can be inspected in detail (this builds on functionality from the
`ReportingTools` package).

```{r}
eaBrowse(comb.res, graph.view=mmu.grn, nr.show=100, html.only = TRUE,
         out.dir = "./reports_1")
```

*Session Info*

```{r sessionInfo}
sessionInfo()
```